---
toc: true
---

<a href="https://colab.research.google.com/github/tuomaseerola/emr/blob/master/Chapter7.ipynb" target="_parent"><img src="https://colab.research.google.com/assets/colab-badge.svg" alt="Open In Colab"/></a>

# Ch. 7 – Inferential statistics

This notebook demonstrates running inferential statistical tests in R.

## Preliminaries

Load libraries and install `MusicScienceData` package where the example data is stored.

```{r}
#| label: libraries
#| warning: false
library(ggplot2,quietly = TRUE)
library(tidyverse,quietly = TRUE)
library(MusicScienceData,quiet=TRUE)
```

## Code 7.1

See text for the explanation.

```{r}
#| label: 7-1
#| warning: false
df <- MusicScienceData::sadness         # define data
t <- t.test(ASM20 ~ gender, data=df)    # t test
print(t$statistic)                      # show the t value

print(scales::pvalue(t$p.value))
dplyr::summarise(dplyr::group_by(df, gender), # means and SDs
                 M=mean(ASM20,na.rm=TRUE),
                 SD=sd(ASM20,na.rm=TRUE))
```

## Code 7.2

```{r}
#| label: 7-2
#| warning: false
df <- MusicScienceData::sadness         # define data
model.aov <- aov(ASM20 ~ age, data=df)  # run anova
F <- summary(model.aov)                 # summarise
print(F)
```

## Code 7.3

```{r}
#| label: 7-3
#| warning: false
TABLE<-TukeyHSD(model.aov,conf.level = 0.95)
print(knitr::kable(TABLE$age,digits = 3,
                   caption = 'Comparison of age groups
                   for Item 20 in ASM survey.',
                   format = 'simple'))
```

## Code 7.4

```{r}
#| label: 7-4
#| warning: false
df <- MusicScienceData::sadness                   # define data
model2.aov <- aov(ASM20 ~ age * gender, data=df)  # run anova
F2 <- summary(model2.aov)
print(F2)
```

## Code 7.5

This analysis requires extra libraries and raw data read from github. The installation might be slow in Colab because of package dependencies.

```{r}
#| label: 7-5
#| eval: false
#| warning: false
#install.packages("lme4",quiet=TRUE)     # Required for LMM analysis
#install.packages("lmerTest",quiet=TRUE) # Optional
library(lme4,quiet=TRUE)
library(lmerTest,quiet=TRUE)
```

```{r}
#| label: load-data
#| eval: true
#| warning: false
library(lme4,quiet=TRUE)
library(lmerTest,quiet=TRUE)
d <- read.csv('https://raw.githubusercontent.com/tuomaseerola/emr/main/data/raw_ratings.csv')
d2 <- dplyr::filter(d,Emotion=='Dimensional')  #
d3 <- dplyr::filter(d2, Category=='Anger' |
  Category=='Fear' |
  Category=='Happy' |
  Category=='Sad' |
  Category=='Tender')
m1 <- lmer(Valence ~ Category * Gender + (1|id) + (1|Track), data = d3)
s <- summary(m1,corr=FALSE)
S<-s$coefficients; S<-round(S,2); S[,5]<-scales::pvalue(S[,5])
print(knitr::kable(S,format = 'simple',
                   caption = 'LMM results of Valence ratings.'))
```

## Code 7.6

```{r}
#| label: code7-6
#| warning: false
d <- read.csv('https://raw.githubusercontent.com/tuomaseerola/emr/main/data/raw_ratings.csv')
S <- d %>%
  filter(Category=='Sad') %>%
  group_by(Category,Gender) %>%
  summarise(M=mean(Valence,na.rm=T),SD=sd(Valence,na.rm=T),
            .groups = 'drop')
print(S)
```

## Code 7.7

```{r}
#| label: 7-7
#| warning: false
library(MusicScienceData)               # loads library w data
gender_age_xtab <- table(MusicScienceData::sadness$age,
                         MusicScienceData::sadness$gender)
print(gender_age_xtab)

result <- chisq.test(gender_age_xtab)   # Chi^2 test
print(result)
```

## Code 7.8

```{r}
#| label: 7-8
#| warning: false
library(MusicScienceData)               # load library w data
data <- MusicScienceData::soundtrack    # define data
r<-cor.test(data$Valence, data$Tension) # calculate correlation
print(r$estimate)                       # print coefficient
##    cor
## -0.827
print(scales::pvalue(r$p.value))        # print pretty p value
## [1] "<0.001"
print(r$parameter)                      # print df
```

## Code 7.9

```{r}
#| label: 7-9
#| warning: false
library(MusicScienceData)               # loads library w data
d1 <- MusicScienceData::soundtrack      # get ratings
d2 <- MusicScienceData::soundtrack_features[,c(2:3,5:6)] # select only some features
d1[,17:21] <- as.data.frame(scale(d2))  # normalise

tmp <- cor(d1[,c(3,17:20)])             # get correlations
print(round(tmp[2:5,1],2))              # display first line
```

## Code 7.10

```{r}
#| label: 7-10
#| warning: false
model.reg <- lm(Energy ~ RMS + sp_centr + spec_rolloff +
  spec_zcr, data = d1)
s <- summary(model.reg) # R2adj = 0.424 (Energy)
print(s)
```

## Code 7.11

```{r}
#| label: 7-11
#| warning: false
r <- cor(d1$Energy, d1$RMS)
print( r^2 )                      # print the squared correlation
summary(lm(Energy ~ RMS,data=d1)) # Summarise regression
```
